Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-26 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
small changes

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

@@

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
% $\ell_{\rm EdGB} \leqslant 1.33$~km
\hs{$\ell_{\rm EdGB} \leqslant 7.1$~km}
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
% $\ell_{\rm EdGB} \leqslant 1.7$~km,
\hs{$\ell_{\rm EdGB} \leqslant 9.1$~km},
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{\hs{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field interaction arising from the original dilatonic coupling, i.e.,
$f(\varphi) \approx \varphi$ in Eq.~\eqref{eq:action_sgb}.
This results in what is often referred to as shift-symmetric scalar-Gauss-Bonnet theory.
In this theory, NSs do not have scalar monopole charge~\cite{Yagi:2015oca}, while BHs do~\cite{Yunes:2011we,Sotiriou:2013qea,Sotiriou:2014pfa}. Finally, we take note constant $\alpha_{\rm EdGB}$ used in Refs.~\cite{Yagi:2011xp,Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} is related to $\ell_{\rm EdGB}$ as $\ell_{\rm EdGB} = 4 \pi^{1/4} |\alpha_{\rm EdGB}|^{1/2}$.
\hscomm{I had to translate between notations and rewrote the bounds quoted for $\alpha_{\rm EdGB}$ in terms of our $\ell_{\rm EdGB}$.}
}}.
Secondly, the scalar field influences the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{align} \label{eq:action_dcs}
    S_{\rm dCS} =
    \int \dV
    \left[
    \hs{\tfrac{1}{16\pi}} R - \tfrac{1}{2}(\partial \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm dCS} \, \vartheta \, {}^{*}RR
    \right],
\end{align}
%
where $\vartheta$ is a pseudoscalar field which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R_{\mu\nu}{}^{\rho\sigma} R^{\nu\mu}{}_{\rho\sigma}$,
%
where ${}^{*}R^{\mu\nu\rho\sigma}$ is the dual of the Riemann tensor
defined as
%
${}^{*}R^{\mu\nu\rho\sigma} =
\epsilon^{\mu\nu}{}_{\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and $\epsilon_{\mu\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The Pontryagin density by itself, $\int \dV {}^{*}RR$, is a boundary
term that does not contribute to the field equations in
four-dimensions~\cite{Jackiw:2003pm}.
%
The Pontryagin density can, however, modify the field equation when coupled to $\vartheta$;
the strength of the coupling set by $\ell_{\rm dCS}$, with dimensions of length.
%
% However, a coupling with $\vartheta$ can modify the field equations.
% %
% The coupling strength is set by the coupling constant $\ell_{\rm dCS}$, with
% dimensions of length.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function on coupling constant $\ell_{\rm EdGB}$ in EdGB gravity
for GW150914 (top panel), GW200129 (middle panel)
and from combining results (bottom panel).
%
In all panels, different line colors correspond to the inclusion ($N_{\rm max} = 1$) or not ($N_{\rm max} = 0$)
of the linear-in-spin QNM correction.
%
The joint posteriors are shown for illustrative purposes only. As we explain in Fig.~\ref{fig:edgb_cdf} and
in the main text, our analysis of these events fails to satisfy the EFT bound~\eqref{eq:eft_bound}. We mark
with vertical lines theh $90\%$ upper credible intervals.}
\label{fig:edgb_exec_sum}
\end{figure}

@@

% --------------------------
\section*{Acknowledgements}
\label{sec:acknowledgements}
%
We thank Emanuele Berti, Andrea Maselli, Caio F. B. Macedo, Deyan Mihaylov,
Serguei Ossokine, Scott E. Perkins, and Helvi Witek for discussions.
%
We are grateful for the computational resources provided by the Max Planck
Institute for Gravitational Physics in Potsdam, specifically the
high-performace computing cluster Hypatia and to Steffen Grunewald for
assistance.
%
The material presented in this manuscript is based upon work supported by NSF’s LIGO Laboratory,
which is a major facility fully funded by the NSF.
%

This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open
Science Center (https://www.gw-openscience.org), a service of LIGO Laboratory, the LIGO Scientific Collaboration
and the Virgo Collaboration, and Zenodo (https://zenodo.org/record/5172704).
%

@@

% \hscomm{There was an error here: I mixed the axial QNMs we used for the nonrotating coefficients, with the polar QNMs
% for which the spin-corrections were computed. Because it is inconsistent, I suggest changing opening paragraph with the highlighted text,
% and then remove the $N_{\rm max} = 1$ results from our EdGB plots. What do you think?}
%
\hs{The spin-corrections to the \emph{polar} gravitational-led modes
were calculated in Ref.~\cite{Pierini:2021jxd} [see, in particular, their Eqs.~(51)
and~(52).]
%
For consistency with our previous discussion, we truncate these equations at
leading-order in $\gamma_{\rm EdGB}$, but we emphasize that we are being inconsistent in
mixing results valid for modes of different parities.
%
We still do so, simply to explore what the rotational corrections to EdGB gravity QNMs \emph{might}
tells us in our ringdown analysis and the results of Ref.~\cite{Pierini:2021jxd} are our best guide
available presently.}
% \abcomm{$\ell_{\rm EdCS}^4$?}.

@@

The QNMs of slowly-rotating BHs in both cubic and quartic EFT of GR where calculated in Ref.~\cite{Cano:2021myl}.
%
For the cubic EFT, we use their Eq.~(67), particularizing it to $\lame = \lamo = 1$.
%
We further linearize that equation in $\chi$ and consider the harmonic $m=2$.
%
This process, shows that the mode labeled ``-'' is the least-damped mode, and it is the one we use.
%
Direct comparison with Eqs.~\eqref{eq:qnm_nongr_iso} results in $p_{\rm cEFT} = 4$,
%
\begin{align}
    \delta \omega_{\rm cEFT}^{(0)} = - 0.5813, \quad \delta \tau_{\rm cEFT}^{(0)} = 2.6469,
    \nonumber \\
    \delta \omega_{\rm cEFT}^{(1)} = -3.8620, \quad \delta \tau_{\rm cEFT}^{(1)} = 265.12.
\end{align}

We proceed in the same away for the quartic EFT. Here we use Eq.~(68) (with
$\epsilon_{1} = 1$ and $\epsilon_{2} = 0$) and Eq.~(70) of
Ref.~\cite{Cano:2021myl},
%
In this case we find that both modes (``-'' and ``+'' in the notation of Ref.~\cite{Cano:2020cao}) make the
fundamental QNM damp fast, and we choose the ``-'' mode which decreases less the damping time. \abcomm{Not clear.
Decreases less with respect to what? Do you mean it has a smaller damping time?}
%
This time we then find that $p_{\rm qEFT} = 6$,
%
\begin{align}
    \delta \omega_{\rm qEFT}^{(0)} = -0.2114, \quad \delta \tau_{\rm qEFT}^{(0)} = -0.6070,
    \nonumber \\
    \delta \omega_{\rm qEFT}^{(1)} = -1.5263, \quad \delta \tau_{\rm qEFT}^{(1)} = 171.35.
\end{align}
%
As with the previous theories, the larger values of some of these parameters are compensated
by the assumptions of weak coupling and small spin used to calculate the QNM frequencies.
\
\section{The estimation of  intrinsic binary parameters in General Relativity and modified theories of gravity}
\label{app:big_contour}

>>>>

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during in the post-merger. So far, none of these null tests have reported any deviation from GR.

@@

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with GW
observations of compact binaries where at least one component is a BH.
%
Indeed, Ref.~\cite{Lyu:2022gdr} placed the bound
%
% $\ell_{\rm EdGB} \leqslant 1.33$~km
\hs{$\ell_{\rm EdGB} \leqslant 7.1$~km}
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
% $\ell_{\rm EdGB} \leqslant 1.7$~km,
\hs{$\ell_{\rm EdGB} \leqslant 9.1$~km},
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{\hs{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field interaction arising from the original dilatonic coupling, i.e.,
$f(\varphi) \approx \varphi$ in Eq.~\eqref{eq:action_sgb}.
This results in what is often referred to as shift-symmetric scalar-Gauss-Bonnet theory.
In this theory, NSs do not have scalar monopole charge~\cite{Yagi:2015oca}, while BHs do~\cite{Yunes:2011we,Sotiriou:2013qea,Sotiriou:2014pfa}. Finally, we take note constant $\alpha_{\rm EdGB}$ used in Refs.~\cite{Yagi:2011xp,Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} is related to $\ell_{\rm EdGB}$ as $\ell_{\rm EdGB} = 4 \pi^{1/4} |\alpha_{\rm EdGB}|^{1/2}$.
\hscomm{I had to translate between notations and rewrote the bounds quoted for $\alpha_{\rm EdGB}$ in terms of our $\ell_{\rm EdGB}$.}
}}.
Secondly, the scalar field influences the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant,
results in a coupling between scalar perturbations and gravitational
perturbations of polar parity which, for instance, breaks the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) of Schwarzschild BHs in GR.

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{align} \label{eq:action_dcs}
    S_{\rm dCS} =
    \int \dV
    \left[
    \hs{\tfrac{1}{16\pi}} R - \tfrac{1}{2}(\partial \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm dCS} \, \vartheta \, {}^{*}RR
    \right],
\end{align}
%
where $\vartheta$ is a pseudoscalar field which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R_{\mu\nu}{}^{\rho\sigma} R^{\nu\mu}{}_{\rho\sigma}$,
%
where ${}^{*}R^{\mu\nu\rho\sigma}$ is the dual of the Riemann tensor
defined as
%
${}^{*}R^{\mu\nu\rho\sigma} =
\epsilon^{\mu\nu}{}_{\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and $\epsilon_{\mu\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The variation of the Pontryagin density is a boundary
term that does not contribute to the field equations in
four-dimensions~\cite{Jackiw:2003pm}.
%
However, the Pontryagin density can modify the field equations when coupled to $\vartheta$;
the strength of the coupling set by $\ell_{\rm dCS}$, with dimensions of length.
%
% However, a coupling with $\vartheta$ can modify the field equations.
% %
% The coupling strength is set by the coupling constant $\ell_{\rm dCS}$, with
% dimensions of length.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function on coupling constant $\ell_{\rm EdGB}$ in EdGB gravity
for GW150914 (top panel), GW200129 (middle panel)
and from combining results (bottom panel).
%
In all panels, different line colors correspond to the inclusion ($N_{\rm max} = 1$) or not ($N_{\rm max} = 0$)
of the linear-in-spin QNM correction.
%
The joint posteriors are shown for illustrative purposes only. As we explain in Fig.~\ref{fig:edgb_cdf} and
in the main text, our analysis of these events fails to satisfy the EFT bound~\eqref{eq:eft_bound}. We mark
with vertical lines the $90\%$ upper credible intervals.}
\label{fig:edgb_exec_sum}
\end{figure}

@@

% --------------------------
\section*{Acknowledgements}
\label{sec:acknowledgements}
%
We thank Emanuele Berti, Andrea Maselli, Caio F. B. Macedo, Deyan Mihaylov,
Serguei Ossokine, Scott E. Perkins, and Helvi Witek for discussions.
%
We are grateful for the computational resources provided by the Max Planck
Institute for Gravitational Physics in Potsdam, specifically the
high-performace computing cluster Hypatia and to Steffen Grunewald for
assistance.
%
The material presented in this manuscript is based upon work supported by NSF’s LIGO Laboratory,
which is a major facility fully funded by the NSF.

This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open
Science Center (https://www.gw-openscience.org), a service of LIGO Laboratory, the LIGO Scientific Collaboration
and the Virgo Collaboration, and Zenodo (https://zenodo.org/record/5172704).

@@

% \hscomm{There was an error here: I mixed the axial QNMs we used for the nonrotating coefficients, with the polar QNMs
% for which the spin-corrections were computed. Because it is inconsistent, I suggest changing opening paragraph with the highlighted text,
% and then remove the $N_{\rm max} = 1$ results from our EdGB plots. What do you think?}
%
\hs{The spin-corrections to the \emph{polar} gravitational-led modes
were calculated in Ref.~\cite{Pierini:2021jxd} [see, in particular, their Eqs.~(51)
and~(52).]
%
For consistency with our previous discussion, we truncate these equations at
leading-order in $\gamma_{\rm EdGB}$, but we emphasize that we are being inconsistent in
mixing results valid for modes of different parities.
%
We still do so, simply to explore what the rotational corrections to EdGB gravity QNMs \emph{might}
tells us in our ringdown analysis and the results of Ref.~\cite{Pierini:2021jxd} are our best guide
presently available.}
% \abcomm{$\ell_{\rm EdCS}^4$?}.

@@

The QNMs of slowly-rotating BHs in both cubic and quartic EFT of GR where calculated in Ref.~\cite{Cano:2021myl}.
%
For the cubic EFT, we use their Eq.~(67), in the particular case of $\lame = \lamo = 1$.
%
%
We then linearize the resulting expression in $\chi$ and consider $m=2$ the harmonic.
%
\hscomm{Instead calling them ``$+$'' and ``$-$'' following the notation used in~\cite{Cano:2021myl},
I call them axial/polar to follow our conventions.}
%
As an outcome, we find that the fundamental axial-parity QNM is the least damped one, and it is
the one we use.
%
Direct comparison with Eqs.~\eqref{eq:qnm_nongr_iso} results in $p_{\rm cEFT} = 4$,
%
\begin{align}
    \delta \omega_{\rm cEFT}^{(0)} = - 0.5813, \quad \delta \tau_{\rm cEFT}^{(0)} = 2.6469,
    \nonumber \\
    \delta \omega_{\rm cEFT}^{(1)} = -3.8620, \quad \delta \tau_{\rm cEFT}^{(1)} = 265.12.
\end{align}

We proceed in the same away for the quartic EFT. Here we use Eq.~(68) (with
$\epsilon_{1} = 1$ and $\epsilon_{2} = 0$) and Eq.~(70) of
Ref.~\cite{Cano:2021myl},
%
In this case we find that both modes (``$-$'' and ``$+$'' in the notation of Ref.~\cite{Cano:2020cao}) make the
fundamental QNM damp fast, and we choose the ``$-$'' mode which decreases less the damping time. \abcomm{Not clear.
Decreases less with respect to what? Do you mean it has a smaller damping time?}
%
This time we then find that $p_{\rm qEFT} = 6$,
%
\begin{align}
    \delta \omega_{\rm qEFT}^{(0)} = -0.2114, \quad \delta \tau_{\rm qEFT}^{(0)} = -0.6070,
    \nonumber \\
    \delta \omega_{\rm qEFT}^{(1)} = -1.5263, \quad \delta \tau_{\rm qEFT}^{(1)} = 171.35.
\end{align}
%
As with the previous theories, the larger values of some of these parameters are compensated
by the assumptions of weak coupling and small spin used to calculate the QNM frequencies.
\
\section{The estimation of  intrinsic binary parameters in General Relativity and modified theories of gravity}
\label{app:big_contour}
